---
title: "Preliminary Analysis"
author: "Kanika Chopra"
date: "2022-11-03"
output: pdf_document
---

# Predicting Parkinson's Based on Auditory Data 


## a) Import Data 
First, we want to read in our data. 

```{r}
library(RCurl) 
data <- read.csv(text = getURL("https://raw.githubusercontent.com/kanikadchopra/Parkinsons-Prediction/main/parkinson_data.csv"))
```

```{r}
attach(data)
```

```{r}
library(tidyverse)
```

```{r}
dim(data)
```

```{r}
glimpse(data)
```

## b) Missing Values
Next, we want to conduct a few quick quality checks, such as checking if there are any missing values. 
```{r}
data %>% summarise_all(~ sum(is.na(.)))
```

We can see that we have zero missing values along all of our columns which is good news as we don't have to handle them in the analysis. 

## c) Preliminary Plots

First, let's look at our predictor variable.
```{r}
barplot(table(status), xlab='Status', ylab='Frequency')
```

Firstly, we can see that our data is skewed in that we have more data on patients with Parkinson's (`status=1`) than patients who do not have Parkinson's (`status=0`).

Next, we look at some of our variables

```{r}
boxplot(status, data$MDVP.Fo.Hz)
boxplot(status, data$MDVP.Fhi.Hz)
boxplot(status, )
```


```{r}
pairs(data[-c(1)])
```


* Histograms, boxplots, scatterplots, comment on anomalies

* Pairs plot
* box plot against the predictor





